---
title: "**Motor Car trends**"
author: "Thej Ratheesh"
format:
  html:
    fontsize: 1.0em
    embed-resources: true
    grid:
      body-width: 800px
  revealjs: 
    slide-number: true
    html-math-method: mathjax
    theme: serif
execute: 
  eval: true
  echo: false
---


## Introduction 

The **`mtcars`** dataset, included in base R, contains data on 32 car models from the 1970s. Each row represents a car, with variables such as horsepower, weight, and miles-per-gallon (MPG). 

In this brief analysis, we explore the relationship between a car's **horsepower** and its **fuel efficiency** measured in MPG.

[**R Documentation Link**](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars).

```{r}
library(knitr)
kable(head(mtcars))

```

## Bar Chart
The bar chart shown displays the Miles per Gallon (MPG) for each car model in the mtcars dataset.

- X-axis: Car models (e.g., Toyota Corolla, Fiat 128, Lincoln Continental).
- Y-axis: Miles per Gallon — a measure of fuel efficiency.
- The bars are sorted in descending order, meaning the most fuel-efficient cars appear on the left, and the least efficient on the right.

```{r}
#| fig-align: center
library(ggplot2)

# Add car names as a column
mtcars$car <- rownames(mtcars)

# Create bar plot
ggplot(mtcars, aes(x = reorder(car, -mpg), y = mpg)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Miles per Gallon (MPG) for Each Car in mtcars Dataset",
       x = "Car",
       y = "Miles per Gallon") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 9))
```

- The Toyota Corolla has the highest MPG (~33.9), making it the most fuel-efficient car in this dataset.
- On the opposite end, the Lincoln Continental has the lowest MPG (~10.4), indicating high fuel consumption.
- There is a clear negative skew, with most cars clustered between 15 and 25 MPG, and only a few reaching above 30 MPG.
- This variation in MPG reflects the impact of factors such as engine size, weight, and horsepower on fuel efficiency.
- Vehicles like Fiat 128 and Honda Civic also perform well, showing that smaller or lighter cars tend to be more efficient.


## MPG Distribution by Cylinder Count:

This graph is a boxplot that visualizes the distribution of miles per gallon (MPG) for cars with different numbers of cylinders (4, 6, and 8) in the mtcars dataset.

Each boxplot shows:

- The median MPG (thick black horizontal line inside each box).
- The interquartile range (IQR) (height of the box — middle 50% of the data).
- The whiskers extend to 1.5×IQR or to the min/max (if no outliers).
- Outliers (individual points outside whiskers) are shown as small circles (e.g., in the 8-cylinder group).


```{r}
#| label: hist-speed
#| fig-align: center
boxplot(mpg ~ cyl, data = mtcars,
        xlab = "Cylinders", ylab = "Miles per Gallon",
        main = "MPG Distribution by Cylinder Count",
        col = c("skyblue", "lightgreen", "orange"))

```


## Pairwise Scatterplot Matrix
This is a pairwise scatterplot matrix. It visualizes the relationships among four variables from the mtcars dataset:

- [mpg: Miles per gallon (fuel efficiency)]{style="color: red;"}
- [hp: Horsepower]{style="color: green;"}
- [wt: Weight (in 1000 lbs)]{style="color: blue;"}
- [qsec: 1/4 mile time (acceleration)]{style="color: purple;"}


```{r}
#| warning: false
#| fig-align: center
library(GGally)
library(ggplot2)

# Subset of mtcars
subset_data <- mtcars[, c("mpg", "hp", "wt", "qsec")]

# Pairwise scatterplot matrix
ggpairs(subset_data,
        title = "Pairwise Scatterplots of mpg, hp, wt, and qsec")


```

- Fuel efficiency (mpg) is most negatively affected by weight (wt) and horsepower (hp)
- Heavier and more powerful cars generally use more fuel.
- The strongest overall relationship in the data is between mpg and wt.


## Summary

From our analysis it can be examined that :

- Fuel efficiency (MPG) varies widely across car models, with compact cars like the Toyota Corolla and Fiat 128 outperforming heavier models like the Lincoln Continental.
- Cars with fewer cylinders (4 or 6) tend to have higher MPG than 8-cylinder models, as seen in the boxplot analysis.
- Pairwise correlations revealed strong negative relationships between MPG and both horsepower (-0.78) and weight (-0.87), suggesting that more powerful and heavier vehicles are less fuel efficient.
- A moderate positive correlation between MPG and qsec indicates that more efficient cars tend to have slightly slower acceleration, potentially due to smaller engines.

lighter and less powerful vehicles tend to be more fuel efficient, which is kind of the expected and practical outcome.
